Towards an Ecological Trait-data Standard

### 2018-10-31 manuscript/manuscript.Rmd
final comments on review, update bibliography

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Malte Jochum \footnotemark[4]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[7]
csl: amnat.csl
fontsize: 11pt
geometry: margin=1.4in
documentclass: scrartcl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

1. Trait-based approaches are widespread throughout ecological research, as they offer great potential to achieve a general understandnig of a wide range of ecological and evolutionary mechanisms. Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
2. We review current initiatives and structures developed for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data. We discuss potential incentives and challenges for the wide adoption of such a standard by data-providers. 
4. The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases and facilitates the creation of interfaces between software tools for trait-data handling and analysis. By aiding decentralised trait-data standardisation, our data standard may facilitate the use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.

@@

@@

Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/)
Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted
Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*
Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g. trait data), that is recognised by a large proportion of members of the application context
Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g. may define a hierarchy of broader or narrower terms
Ontology | *Controlled vocabulary* that defines *concepts* in relation to each other in a formal language, e.g. provides cross-references and semantic meaning; a semantic model of the objects and their relationships in a domain of interest [@gruber95]; commonly published in OWL format for machine readability.
Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e. specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/)
Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries
Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time
Database | A suite of data compiled from multiple *datasets*, i.e. from multiple study contexts or observation types; may take the form of a single two-dimensional *data tables*, but mostly is organised as multiple data tables linked via *identifiers* into *relational databases*; may be built using a specific database software or language (e.g. SQL)
Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs)
Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally 
Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *relational database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*  
Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://t-sita.cesab.org/BETSI_vizInfo.jsp?trait=Body_length'
Web Service | A service to access *online databases* directly and computationally, i.e. by calls from a software tool; returns content of the database according to the request in different formats and *data standards*
Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g. client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider
Online Portal | A website designed as a platform for the exchange of information, e.g. trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access
File repository | A short-term storage of datasets or long-term archiving on *file-hosting services*; online repositories make data available for public access, provide *metadata* and (not always) facilitate citations via persistent identifiers, e.g. DOIs (Digital Object Identifiers) 
File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org
Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97]
Darwin Core Standard (DwC) | Body  of *terminologies* providing terms intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/)
Darwin Core Archive (DwC-A) | A file archive, or *repository* structure, that contains *metadata* (specified e.g. using Ecological Metadata Language, EML) and primary data combined into a *relational database* via *identifier* columns [@gbif17,  http://tools.gbif.org/dwca-assistant/]
Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures; Example: Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits @moretti17
Occurrence | The observation context of a single entity of a taxon, i.e.  an individual observed at a particular place and time (http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context

@@

<!--
Measurement |  
Aggregated Value |
Value | 
Fact | 
-->

@@

Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
Recent years have seen a proliferation of trait-based research in a wide range of fields and have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10]. This research provides the mechanistic framework for linking climate change or anthropogenic land use to biodiversity and its related functions [@diaz11; @lavorel12; @allan15].
Species traits have been suggested as indivator variable for monitoring  ecosystem health at the individual level [@kissling18].  
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that most species are yet undescribed and little is known of their interactions with other organisms and their environment. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

@@

<!-- 2. trait databases have been developed, with standard structure and terms --> 
Despite the importance of functional trait data, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa as well as in evolutionary deep-time. 
However, the heterogeneity of data arising from different research contexts render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
<!-- i.e. differences in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements--> 
To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centred around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. Less well-studied taxa and specialised research questions lack the resources for such an endeavour.
Besides initiatives aiming at assembling data, tools to enable the compatibility of data across databases are being developed. These include software to access trait data from the internet [e.g. @chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of standard terms [@walls12; @garnier17].

<!-- 4. distributed trait data must be integrated --> 
Meanwhile, open and reproducible science has become mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding agencies [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15]. 
As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services (e.g. Figshare.com, see Table 1), where no data standards areenforced upon the uploaded material [@wilkinson16]. 
It is thus likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient reuse and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, and initiatives for data standardisation from the pragmatic view of data providers, data curators and data users, as well as database managers. We discuss current efforts to making trait data visible, accessible, interoperable and re-useable in downstream data analysis, as demanded by the FAIR guiding principles for scientific data [@wilkinson16]. 
Furthermore, we show how the current deficit in the standardisation of primary data hampers the implementation of interoperability and reuse of trait data. 
Based on these considerations, we propose a versatile vocabulary for describing trait datasets that builds upon, and is compatible with , existing terminology standards for biodiversity data. 
Since this standard vocabulary relies on the adoption of a broad research community, we discuss incentives for its use and lay out mechanisms for a future consensus-building and community development towards an accessible ecological trait data standard.
<!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
<!-- FDS: Why 'ecological' trait data? what is special about this subject? Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->

@@

The need for standardising trait-data arises from the prospective gain of compiling heterogeneous trait-datasets for data synthesis. 
Often, the scientific scope and focus differs between data providers measuring and assessing the trait data in the first place and data users who re-use published data for a broader synthesis application. Data curators are taking up the task of providing compiled and harmonised datasets or databases. Data managers are concerned with the development of complex digital infrastructures for handling and analysing large amounts of data. 
These idealised roles of researchers cooperatively are dealing with trait-data standardisation throughout the data life-cycle. In this chapter, we review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):

1. Initiatives that provide trait *datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature.
2. Initiatives that aim to harmonise trait data from the literature or from direct measurements into trait *databases* and make those data widely available. 
3. Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits and provide standard *terminologies* in the form of *thesauri* and *ontologies*.
4. Initiatives that aim to leverage *relational database* structures and *semantic web* technology to link trait data to a wider set of biodiversity data.

We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17]. 
We show how the degree of trait-data standardisation in existing datasets is highly variable and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources. 
The objective of this review is to raise awareness of the generic structure of trait data and aid researchers in how to share and publish their own datasets in an appropriate form.

@@

In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g. reproductive traits, feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published alongside scientific articles, or as standalone data publications [see @kleyer08 for a review on plant data; for animal data, see e.g., @gossner15; see Supplementary Material A, Table A1].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->

Today, the online publication of such data is greatly facilitated by *file hosting services* (e.g. Figshare.com, see Table 1), which warrant long-term accessibility and citeability via DOIs, and govern data sharing via licensing.
These platforms offer publicly accessible *repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management. 
However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make extraction tedious and dependent on commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation. As a result, published datasets greatly vary in structure, nomenclature and metadata quality.

For trait data, there are common issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives. Furthermore, metadata varies in its level of detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this intuitive *data table*, each row contains a species (or taxon) for which multiple traits are reported in columns. <!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).  -->
Similarly, when reporting raw data, researchers place observations of individual organisms (i.e. the *occurrences*) in rows with multiple trait measurements applied to the same individual across multiple columns. 
Co-variates on the taxon, the individual specimen (e.g. sex or lifestage) or context of observation (e.g. time and place of sampling) would be placed in additional columns, and would further expand the two-dimensional data table. The resolution or scope of these co-variates varies greatly depending on the research question and observation context. 
Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14]. A task that is taken up by data curators, who compile heterogeneous data into comprehensive databases (see next section).

@@

## Database initiatives

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @kleyer08; @oliveira17, see Supplementary Material A, Table A1].
While mostly concerned with issues of heterogeneity in units or factor levels, and aiming for high taxonomic coverage, few of these *databases* apply a standardised terminology for taxa or traits that would allow them to be efficiently linked to other databases.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus. 
Just as the individual datasets described above, many of these databases are published as static data tables on journal websites or open-access file hosting platforms, but they may be updated irregularly.

As they deal with much larger amounts of data, initiatives that form around natural history museum collections are traditionally more concerned with standardisation. 
The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g. www.markmybird.org), 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

<!-- database backend, selective download, user access management --> 
Specialised *online portals* have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform. The *online database* enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
For animal trait data, however, a single unified portal and harmonising scheme is still lacking. The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].

Nonetheless, initiatives for particular groups of animals do exist: examples are the BETSI database on soil invertebrate traits [http://betsi.cesab.org/; @pey14], the Carabids.org web portal  (http://www.carabids.org/), or the Coral Trait Database [@madin16, see Supplementary Material A, Table A1]. 
The role of online portals and database initiatives in standardising data and making them more accessible is paramount. 
Trait-data portals incentivise data submissions by offering increased data visibility and usage, while providing data-use policies that secure author attribution and, potentially, co-authorship of associated articles. However, maintaining centralised database infrastructures is costly and requires long-term funding [@bach12].

## Thesauri and Ontologies for traits

@@

While trait thesauri and ontologies typically define traits for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1]  \footnotemark[2]
- name: David Fichtmüller \footnotemark[5]
- name: Martin M. Gossner \footnotemark[8]
- name: Anton Güntsch \footnotemark[5]
- name: Malte Jochum \footnotemark[4]
- name: Birgitta König-Ries\footnotemark[6]
- name: Gaëtane Le Provost \footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[6]
- name: Caterina Penone \footnotemark[4]
- name: Nadja K. Simons \footnotemark[7]
csl: mee.csl
fontsize: 11pt
geometry: margin=1.4in
documentclass: scrartcl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

1. Trait-based approaches are widespread throughout ecological research, as they offer great potential to achieve a general understandnig of a wide range of ecological and evolutionary mechanisms. Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
2. We review current initiatives and structures developed for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data. We discuss potential incentives and challenges for the wide adoption of such a standard by data providers. 
4. The use of a standard vocabulary allows trait datasets from heterogeneous sources to be compiled more easily into databases and facilitates the creation of interfaces between software tools for trait-data handling and analysis. By aiding decentralised trait-data standardisation, our data standard may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.

@@

Application Programming Interface (API) |  A set of clearly defined methods of communication between  software components, e.g. client software and the *web service* of an *online database*; APIs are usually documented on the website of the service provider.
Concept | An idea, notion or object that is made explicit in an information context by a *term* definition, and referenced to a *URI* or other accessible reference (https://www.w3.org/TR/skos-reference/).

@@

Controlled vocabulary | A list of *terms* that gives all valid consensus terms for a prarticular context, while no unlisted entries are accepted.
Darwin Core Archive (DwC-A) | A file archive that contains primary data tables, linked via *identifier* columns, and *metadata* (specified e.g. using Ecological Metadata Language, EML) [@gbif17,  http://tools.gbif.org/dwca-assistant/].
Darwin Core Standard (DwC) | Body of *terms* intended to facilitate the sharing of information about biological diversity (http://rs.tdwg.org/dwc/).
Dataset | A set of measurements and observations, often stored in a data-table and originating from a single experimental set-up or study context; can be considered as being internally homogeneous across all data entries.
File repository | A storage or archiving of datasets on file-hosting services like Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org; online repositories make data available for public access, provide *metadata*, state conditions of re-use, and (not always) facilitate citations via persistent identifiers, e.g. DOIs (Digital Object Identifiers). 
Identifier (ID) | A unique label that relates row entries within and across *datasets*; may be used to connect multiple *data tables* into a *relational database*; can be user-specific or, in form of a *URI*, point to a globally valid *ontology* or *thesaurus*  .
Metadata | Data documentation of the higher level information or instructions; describe the content, context, quality, structure, provenance and accessibility of a data object [@michener97].
Occurrence | The observation context of a single entity of a taxon, i.e. data on an individual observed at a particular place and time. (http://rs.tdwg.org/dwc/terms/Organism); Sometimes used as synonym of 'observation' in data management context.
Online database | A *database* that is made accessible on the internet; offering forms for filtering and downloading subsets of the data; some online databases offer access via a *webservice* and an *API* that can be addressed computationally.
Ontology | a semantic model of the objects and their relationships in a domain of interest [@gruber95]; defines *terms* and *concepts* in a formal language that provides cross-references and semantic meaning; commonly published in OWL format for machine readability.
Relational database | Two or more *data tables* that are related by common information contained in one or more columns; common information is usually labelled by *identifiers* (IDs).
Semantic web | An extension of the World Wide Web that aims for machine-readable meaning of information via well-defined *data standards*, *ontologies* and exchange protocols [@berners-lee01]; the World Wide Web Consortium (W3C) defines standards, i.e. specifications of protocols and technologies for the semantic web (http://www.w3.org/standards/semanticweb/).
Terminology | The  body of *terms* and *concepts* used with a particular application in a subject of study, usually formalised in a *thesaurus* or *ontology*.
Thesaurus | *Controlled vocabulary* that provides key *terms* with their associated *concepts* and relations for a specific field or domain of interest [@laporte13]; e.g. may define a hierarchy of broader or narrower terms.
Uniform Resource Identifier (URI) | An unambiguous pointer to a unique resource on the internet; used to refer to single terms of a *thesaurus* or *ontology*; Example: 'http://purl.obolibrary.org/obo/TO_0000391'.
Web Service | A service to access *online databases* directly and computationally, i.e. by calls from a software tool; returns content of the database according to the request in different formats and *data standards*.

@@

<!--Data table | A two-dimensional spread-sheet containing data organised in rows and columns; in most cases these data are considered 'static', i.e. they are not altered or filtered across time.-->
<!--Database | A suite of data compiled from multiple *datasets*, i.e. from multiple study contexts or observation types; may take the form of a single two-dimensional *data tables*, but mostly is organised as multiple data tables linked via *identifiers* into *relational databases*; may be built using a specific database software or language (e.g. SQL).-->
<!--Data standard | A published set of instructions and *terminologies* for storing and exchanging data content of a particular type (e.g. trait data), that is recognised by a large proportion of members of the application context.-->
<!--File-hosting service | An online platform that hosts *datasets* or entire *file repositories* and provides access to a wide audience on the internet; Examples: Figshare.com, Dryad (datadryad.org), Researchgate.net, or Zenodo.org.-->
<!--Method handbook | A listing of consensus methodology that is to be applied to acquire a particular measure, thus formalising the precise *concepts* of measures; Example: Handbook of protocols for standardized measurement of terrestrial invertebrate functional traits [@moretti17].-->
<!--Online portal | A website designed as a platform for the exchange of information, e.g. trait data; a portal may include a communication forum, data upload forms, a database access point, and advanced user management for data access.-->
<!--
Measurement |  
Aggregated Value |
Value | 
Fact | 
-->

@@

Functional traits are phenotypic (i.e. morphological, physiological, behavioral) characteristics that are related to the fitness and performance of an organism [@mcgill06; @violle07]. 
Recent years have seen a proliferation of trait-based research in a wide range of fields: trait data have been used to understand the evolutionary basis of individual-level properties [@salguero-gomez16], global patterns of biodiversity [@diaz16] and the relationship between ecosystem functions and the functional composition of species assemblages [@mouillot13; @perovic15; @bello10]. This research provides the mechanistic framework for linking climate change or anthropogenic land use to biodiversity and its related functions [@diaz11; @lavorel12; @allan15].
Species traits have been suggested as indicator variables for monitoring  ecosystem health at the individual level, like for instance changes in average body sizes in a population of fish [@kissling18]. 
 <!-- ;@wood15;  @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15-->
Using traits is also a promising means of bypassing taxonomic impediment, i.e. the fact that most species are yet undescribed and little is known of their interactions with other organisms and their environment. This is because functional traits allow us to infer the ecological role of organisms  from their apparent features, regardless of their taxonomic identity [@leprovost17; @duarte11; @schrodt15].

@@

<!-- 2. trait databases have been developed, with standard structure and terms --> 
Despite the importance of functional trait data, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and compatibility of trait data to achieve sufficient taxonomic and regional coverage, both of present-day taxa as well as in evolutionary deep-time. 
However, the heterogeneity of data arising from different research contexts render trait data extremely heterogeneous and make the task of data compilation time-consuming and error-prone.
<!-- i.e. differences in taxonomic nomenclature and resolution (e.g. reported on species level or aggregated on higher taxonomical orders), the scale and place of the study context, or the accurracy of the methodology applied in measurements--> 
To date, trait data have traditionally been harmonised and compiled into centralised databases only for specific organism groups and regional scope, often centered around particular research questions [e.g. PanTHERIA, @jones09; TRY, @kattge11; AmphiBio, @oliveira17]. Less well-studied taxa and specialised research questions lack the resources for such an endeavour.
Besides initiatives aiming at assembling data, tools to enable the compatibility of data across databases are being developed. These include software to access trait data from the internet [e.g. @chamberlain17; @ankenbrand18], semantic-web standards [@wieczorek12; @page08] and thesauri of consensus terms [@walls12; @garnier17].

<!-- 4. distributed trait data must be integrated --> 
Meanwhile, open and reproducible science has become mainstream: publication of research data without access restrictions, with structured metadata and in accordance to data standards to enable their reuse, has become the declared goal of an open biodiversity knowledge management (http://www.bouchoutdeclaration.org/) and is increasingly demanded by journals and public research funding agencies [@emerson15; @allianceofgermanscienceorganisations10; @royalsocietysciencepolicycentre12; @swan12; @allison15]. 
As a result, an increasing number of individual research projects publish their primary data on general-purpose file hosting services, where no data standards are enforced upon the uploaded material [@wilkinson16]. 
It is thus likely that trait data will become increasingly available, but a lack of data and metadata standardisation will hamper the efficient reuse and synthesis of published datasets.

In this paper, we review existing trait databases and online portals, and initiatives for data standardisation from the pragmatic view of data providers, data curators and data users, as well as database managers. We discuss current efforts to make trait data visible, accessible, interoperable and re-useable in downstream data analysis, as demanded by the FAIR guiding principles for scientific data [@wilkinson16]. 
Furthermore, we show how the current deficit in the standardisation of primary data hampers the implementation of interoperability and reuse of trait data. 
Based on these considerations, we propose a versatile vocabulary for describing trait datasets that builds upon, and is compatible with, existing terminology standards for biodiversity data. 
Since this standard vocabulary relies on the adoption by a broad research community, we discuss incentives for its use and lay out mechanisms for future consensus-building and community development towards an accessible ecological trait data standard.
<!--Finally, we present an R package that assists the harmonisation of trait data from distributed sources. -->
<!-- FDS: Why 'ecological' trait data? what is special about this subject? Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->

@@

The need for standardising trait-data arises from the prospective gain of compiling heterogeneous trait-datasets for data synthesis. 
Often, the scientific scope and focus differs between data providers measuring and assessing the trait data in the first place and data users who re-use published data for a broader synthesis application. Furthermore, data curators and data managers are taking up the task of providing compiled and harmonised data and prepare them for future use and long-term preservation. Data managers are concerned with the development of complex digital infrastructures for handling and analysing large amounts of data.  
These are idealised roles of researchers that are dealing with trait-data standardisation throughout the data life-cycle. In this chapter, we review four types of initiatives that are of relevance for trait-data standardisation (see Glossary in Table 1 for italicised terms):

1. Initiatives that provide trait *datasets* which have been assembled out of a particular research interest, either by measurement or collated from the literature.
2. Initiatives that aim to harmonise trait data from the literature or from direct measurements into database infrastructures and make those data widely available on the Internet. 
3. Initiatives that aim at the standardisation and development of consensus measurement methods and definitions for traits and provide standard *terminologies*.
4. Initiatives that aim to combine data (1 & 2) and terminologies (3) to link trait data to a wider set of biodiversity data.

We consider these initiatives separately although they are often developed in conjunction to serve a particular database project, such as the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17]. 
We show how the degree of trait-data standardisation in existing datasets is highly variable, and which tools and standards are currently applied to achieve harmonisation of data from multiple, distributed sources. 
The objective of this review is to raise awareness of the generic structure of trait data and aid researchers in how to share and publish their own datasets in an appropriate form.

@@

In the field of comparative biology, morphological traits, such as traits  related to flower shape, leaf and stem structures for plants or wing and beak measurements for birds, as well as life-history traits such as Ellenberg values for plants or physiological and reproductive traits for animals (e.g. feeding biology, dispersal, metabolic rates and body size) have been assessed for decades and have been published in regular journal articles or books. 
With the rise of ecological trait-based research, measurements and information available from species descriptions have been compiled into project-specific datasets that typically comprise a local set of taxa and a focal set of traits. 
A plethora of such static datasets has been published alongside scientific articles, or as standalone data publications [see @kleyer08 for a review on plant data; for animal data, see e.g., @gossner15; see Supplementary Material A, Table A1].<!--  @sokol11;  @renner14; @vanderplas17; @price13; -->

Today, the online publication of such data is greatly facilitated by file hosting services (e.g. Figshare, Zenodo, Researchgate, Data Dryad), which warrant long-term accessibility, and citeability via DOIs, and govern data sharing via license statements.
These platforms offer the hosting of publicly accessible *file repositories* at low-cost or for free, which makes them attractive for small and intermediate-sized research projects that cannot dedicate extra resources for data management. 
However, although open for download, the trait datasets on data repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make extraction tedious and dependent on specific commercial software, putting the long-term and open accessibility of these data at risk. 
Most importantly, these platforms enable public hosting of data with very low quality thresholds regarding *metadata* documentation and data standardisation. As a result, published datasets greatly vary in structure, nomenclature and metadata quality.

For trait data, there are common issues arising from the variability of data structures. 
For instance, the column descriptions and terminology applied to taxa and traits are mostly project specific and rarely chosen to allow translation into larger database initiatives. Furthermore, the metadata vary in their level of detail, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
In terms of structure, trait data usually are reported in a species$\times$traits wide-table format. In this intuitive data table, each row represents a species (or taxon) for which multiple traits are reported in columns. <!-- Also, sources of  measurement bias are documented (methodology, observer ID, n of observations).  -->
Similarly, when reporting raw data, researchers place observations of individual organisms (i.e. the *occurrences*) in rows with multiple trait measurements applied to the same individual across multiple columns. 
Co-variates on the taxon, the individual specimen (e.g. sex or lifestage) or context of observation (e.g. time and place of sampling) would be placed in additional columns, and would further expand the two-dimensional data table. The resolution or scope of these co-variates varies greatly depending on the research question and observation context. 
Variability in the number and meaning of columns in these data tables requires tedious manual adjustments when merging multiple datasets [@wickham14]. A task that is taken up by data curators, who compile heterogeneous data into comprehensive databases (see next section).

@@

## Data aggregation initiatives

<!-- community database backend, but static publication --> 
In the past two decades, many distributed trait datasets have been aggregated and harmonised into greater collections with particular taxonomic or regional focus [e.g. @kleyer08; @oliveira17, see Supplementary Material A, Table A1].
While mostly concerned with issues of heterogeneity in units or levels of categorical variables, and aiming for high taxonomic coverage, few of these datasets apply a standardised terminology for taxa or traits that would allow them to be efficiently linked to other datasets.<!--and aiming for taxonomic completeness, the products of these initiatives employ a project specific structure and vocabulary. -->
Documentation of metadata and methodology differs in the level of detail, depending on the research focus. 
Trait data compilations are often labelled 'database', although they do not formally provide data in a database structure in the strict sense of biodiveristy informatics.  
Instead, the data are released as static data tables of raw measurements or aggregate trait values on journal websites or open-access file hosting platforms, which may be updated irregularly.

As they deal with much larger amounts of data, initiatives that compile data from natural history museum collections are traditionally more concerned with standardisation. 
The amount of morphological measurements data extracted from museum collections and herbaria is likely to skyrocket in the near future due to digitisation efforts [@vollmar10; @blagoderov12] supported by new technology for scanning and pattern recognition [@younis18; @mantle12; @strobel18] and citizen science initiatives (e.g. www.markmybird.org). 
For example, the VertNet database compiled and harmonised large quantities of vertebrate trait data from collections [@guralnick16]. The resulting data are published as versioned data tables which are updated as new data sources become available (vertnet.org).

<!-- database backend, selective download, user access management --> 
Specialised online portals have been created to attract data submissions from a defined research field and take care of data harmonisation, thereby greatly facilitating data synthesis.
<!--For example, traits on crop plants have been collected in agronomy databases like  BetyDB [https://www.betydb.org/; @lebauer17].-->
For example, by aiming for a universal framework for plant traits, the TRY database [@kattge11]  attracted more data submissions and downloads than any other trait-data platform. The online portal enables selective data download and user permission and rights management. As a community effort, TRY serves as a network for consensus building on trait definitions [@garnier17] and measurement methodology [@perez-harguindeguy13] (see next section). 
For animal trait data, however, a single unified platform and harmonising scheme is still lacking. The fact that animals span multiple trophic levels and possess diverse body structures makes this a more complex task than for plants [@moretti17].

Nonetheless, initiatives for particular groups of animals do exist: examples are the BETSI database on soil invertebrate traits [http://betsi.cesab.org/; @pey14], the Carabids.org web portal  (http://www.carabids.org/), the Coral Trait Database [@madin16], or the Global Ants Database [@parr17, see Supplementary Material A, Table A1]. 
The role of online portals and database initiatives in standardising data and making them more accessible is paramount. 
Trait-data portals incentivise data submissions by offering increased data visibility and usage, while providing data-use policies that secure author attribution and, potentially, co-authorship of associated articles. However, maintaining centralised database infrastructures is costly and requires long-term funding [@bach12].

## Terminology standards for traits

@@

While trait thesauri and trait ontologies typically define concepts of measurements and observations for focal groups of organisms, they do not specify the  format or structure in which trait data should be stored and linked to further standard terminologies, such as standard taxonomy nomenclatures.
